quollr: An R Package for Visualizing 2D Models from Nonlinear Dimension Reductions in High Dimensional Space

### 2024-04-01 paper-quollr.Rmd
Add prediction plots for training and test data

```{r load-libraries}
library(quollr)
library(tibble)
library(knitr)
library(kableExtra)
library(ggplot2)
library(dplyr)

@@

@@

@@

@@

@@

@@

`predict_emb()` function is used to predict 2D embedding for a new data point using the fitted model. This function is useful to predict 2D embedding irrespective of the NLDR technique.

>>>>

```{r load-libraries}
library(quollr)
library(tibble)
library(knitr)
library(kableExtra)
library(ggplot2)
library(dplyr)
library(ggbeeswarm)
library(patchwork)

@@

```{r}

## Decide by looking at MSE plot
num_bins_x_s_curve <- 8
num_bins_y_s_curve <- 14
hex_size_s_curve <- 0.1
hb_obj_s_curve <- hex_binning(data = s_curve_noise_umap_scaled, x = "UMAP1", 
                      y = "UMAP2", num_bins_x = num_bins_x_s_curve, 
                      num_bins_y = num_bins_y_s_curve, x_start = NA, y_start = NA, 
                      buffer_x = NA, buffer_y = NA, hex_size = hex_size_s_curve, col_start = "UMAP")
## Data set with all possible centroids in the hexagonal grid
all_centroids_df <- as.data.frame(do.call(cbind, hb_obj_s_curve$centroids))
## Generate all coordinates of hexagons
hex_grid <- as.data.frame(do.call(cbind, hb_obj_s_curve$hex_poly))
## To obtain the standardise counts within hexbins
counts_df <- as.data.frame(do.call(cbind, hb_obj_s_curve$std_cts))
df_bin_centroids <- extract_hexbin_centroids(centroids_df = all_centroids_df, 
                                             counts_df = counts_df)
## remove low-density hexagons
hex_grid_with_counts <- dplyr::left_join(hex_grid, counts_df, by = c("hex_poly_id" = "hb_id"))

centroid_plot <- ggplot(data = all_centroids_df, aes(x = c_x, y = c_y)) +
  geom_point(color = "#33a02c") +
  coord_fixed() +
  theme_void() +
  theme(plot.title = element_text(size = 7, hjust = 0.5,
vjust = -0.5),
        axis.title.x = element_blank(), axis.title.y = element_blank(),
        axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        axis.text.y = element_blank(), axis.ticks.y = element_blank(),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank(), #change legend key width
        legend.title = element_text(size=8), #change legend title font size
        legend.text = element_text(size=6))

full_grid <- ggplot(data = hex_grid, aes(x = x, y = y)) +
  geom_polygon(fill = "#ffffff", color = "black", aes(group = hex_poly_id)) +
  geom_point(data = all_centroids_df, aes(x = c_x, y = c_y), color = "#33a02c") +
  coord_fixed() +
  theme_void() +
  theme(legend.position="none", legend.direction="horizontal", plot.title = element_text(size = 7, hjust = 0.5,
vjust = -0.5),
        axis.title.x = element_blank(), axis.title.y = element_blank(),
        axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        axis.text.y = element_blank(), axis.ticks.y = element_blank(),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank(), #change legend key width
        legend.title = element_text(size=8), #change legend title font size
        legend.text = element_text(size=6))

std_pts_grid <-  ggplot(data = hex_grid_with_counts, aes(x = x, y = y)) +
  geom_polygon(color = "black", aes(group = hex_poly_id, fill = std_counts)) +
  geom_point(data = s_curve_noise_umap_scaled, aes(x = UMAP1, y = UMAP2), color = "black", alpha = 1) +
  scale_fill_viridis_c(direction = -1, na.value = "#ffffff", option = "C") +
  coord_fixed() +
  theme_void() +
  theme(legend.position="none", legend.direction="horizontal", plot.title = element_text(size = 7, hjust = 0.5,
vjust = -0.5),
        axis.title.x = element_blank(), axis.title.y = element_blank(),
        axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        axis.text.y = element_blank(), axis.ticks.y = element_blank(),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank(), #change legend key width
        legend.title = element_text(size=8), #change legend title font size
        legend.text = element_text(size=6))

umap_data_with_hb_id <- as.data.frame(do.call(cbind, hb_obj_s_curve$data_hb_id))

model_object <- fit_highd_model( training_data = s_curve_noise_training, 
                                 nldr_df_with_id = s_curve_noise_umap_scaled, 
                                 x = "UMAP1", y = "UMAP2", 
                                 num_bins_x = num_bins_x_s_curve, 
                                 num_bins_y = num_bins_y_s_curve, 
                                 x_start = NA, y_start = NA, 
                                 buffer_x = NA, buffer_y = NA, 
                                 hex_size = hex_size_s_curve,
                                 is_rm_lwd_hex = FALSE, 
                                 benchmark_to_rm_lwd_hex = NA, 
                                 col_start_2d = "UMAP", 
                                 col_start_highd = "x")
df_bin_centroids_s_curve <- model_object$df_bin_centroids
df_bin_s_curve <- model_object$df_bin

## Triangulate bin centroids
tr1_object_s_curve <- tri_bin_centroids(df_bin_centroids_s_curve, x = "c_x", y = "c_y")
tr_from_to_df_s_curve <- gen_edges(tri_object = tr1_object_s_curve)
## Compute 2D distances
distance_s_curve <- cal_2d_dist(tr_coord_df = tr_from_to_df_s_curve, 
                             start_x = "x_from", start_y = "y_from", 
                             end_x = "x_to", end_y = "y_to", 
                             select_vars = c("from", "to", "distance"))

## To find the benchmark value
benchmark_s_curve <- find_lg_benchmark(distance_edges = distance_s_curve, distance_col = "distance")
benchmark_s_curve <- 0.5

trimesh_s_curve <- ggplot(df_bin_centroids_s_curve, aes(x = c_x, y = c_y)) +
  geom_trimesh() +
  theme_linedraw() +
  theme(plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
        axis.title.x = element_blank(), axis.title.y = element_blank(),
        axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        axis.text.y = element_blank(), axis.ticks.y = element_blank(),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  annotate(geom = 'text', label = 'a', x = -Inf, y = Inf, hjust = -0.5, vjust = 1.5, size = 3) +
  labs(colour = "")

trimesh_colored_s_curve_umap <- vis_lg_mesh(distance_edges = distance_s_curve, benchmark_value = benchmark_s_curve,
tr_coord_df = tr_from_to_df_s_curve, distance_col = "distance")

trimesh_colored_s_curve_umap <- trimesh_colored_s_curve_umap +
  theme_linedraw() +
  theme(legend.position = "bottom", plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
        axis.title.x = element_blank(), axis.title.y = element_blank(),
        axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        axis.text.y = element_blank(), axis.ticks.y = element_blank(),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  annotate(geom = 'text', label = 'b', x = -Inf, y = Inf, hjust = -0.5, vjust = 1.5, size = 3) +
  labs(colour = "")

trimesh_removed_s_curve_umap <- vis_rmlg_mesh(distance_edges = distance_s_curve, benchmark_value =
benchmark_s_curve, tr_coord_df = tr_from_to_df_s_curve, distance_col = "distance")
trimesh_removed_s_curve_umap <- trimesh_removed_s_curve_umap +
  theme_linedraw() +
  theme(plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
        axis.title.x = element_blank(), axis.title.y = element_blank(),
        axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        axis.text.y = element_blank(), axis.ticks.y = element_blank(),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  annotate(geom = 'text', label = 'c', x = -Inf, y = Inf, hjust = -0.5, vjust = 1.5, size = 3)

df_all_s_curve <- dplyr::bind_cols(s_curve_noise_training |> dplyr::select(-ID), umap_data_with_hb_id)

# show_langevitour(df_all_s_curve, df_bin_s_curve, df_bin_centroids_s_curve, benchmark_s_curve, distance_s_curve, "distance", col_start = "x")

### Define type column
df <- df_all_s_curve |>
  dplyr::select(tidyselect::starts_with("x")) |>
  dplyr::mutate(type = "data") ## original dataset

df_b <- df_bin_s_curve |>
  dplyr::filter(hb_id %in% df_bin_centroids_s_curve$hexID) |>
  dplyr::mutate(type = "model") ## Data with summarized mean
## Reorder the rows of df_b according to the hexID order in df_b_with_center_data
df_b <- df_b[match(df_bin_centroids_s_curve$hexID, df_b$hb_id),] |>
  dplyr::select(-hb_id)
df_exe <- dplyr::bind_rows(df_b, df)
## Set the maximum difference as the criteria
distance_df_small_edges <- distance_s_curve |>
  dplyr::filter(distance < benchmark_s_curve)
## Since erase brushing is considerd.
tour <- langevitour::langevitour(df_exe[1:(length(df_exe)-1)],
                         lineFrom = distance_df_small_edges$from,
                         lineTo = distance_df_small_edges$to,
                         group = df_exe$type, pointSize = append(rep(0, NROW(df_b)), rep(1, NROW(df))),
                         levelColors = c("#6a3d9a", "#33a02c"))
```

@@

```{r, out.width="100%", fig.cap = "Illustration of key steps of the hexagonal bining algorithm: (a) first, all hexagonal bin centroids are created; (b) second, full hexagon grid is created; and (c) finally, assign the data to hexagons. The centroids are used to generate hexagonal coordinates. Once the full hexagonal grid is generated, by computing the 2D euclidean distance from each NLDR points to hexagonal bin centroids, the nearest NLDR points are allocated to the nearest hexagoanl bin."}

centroid_plot + full_grid + std_pts_grid +
  plot_annotation(tag_levels = 'a') +
  plot_layout(ncol = 3) &
  theme(plot.tag = element_text(size = 8))
```

@@

`predict_emb()` function is used to predict 2D embedding for a new $p$-D data point using the fitted model. This function is useful to predict 2D embedding irrespective of the NLDR technique.

```{r}
## Prediction for training data
pred_emb_list_training <- predict_emb(test_data = s_curve_noise_training, 
                             df_bin_centroids = df_bin_centroids_s_curve, 
                             df_bin = df_bin_s_curve, type_NLDR = "UMAP")
pred_df_training <- as.data.frame(do.call(cbind, pred_emb_list_training))

pred_points_training <- ggplot(data = s_curve_noise_umap_scaled, aes(x = UMAP1, y = UMAP2)) +
  geom_point() +
  geom_point(data = pred_df_training, aes(x = pred_UMAP_1, y = pred_UMAP_2), colour = "#de2d26") +
  annotate(geom = 'text', label = 'a', x = -Inf, y = Inf, hjust = -0.5, vjust = 1.5, size = 3)

## Prediction for test data
pred_emb_list_test <- predict_emb(test_data = s_curve_noise_test, 
                             df_bin_centroids = df_bin_centroids_s_curve, 
                             df_bin = df_bin_s_curve, type_NLDR = "UMAP")
pred_df_test <- as.data.frame(do.call(cbind, pred_emb_list_test))

pred_points_test <- ggplot(data = s_curve_noise_umap_scaled, aes(x = UMAP1, y = UMAP2)) +
  geom_point() +
  geom_point(data = pred_df_test, aes(x = pred_UMAP_1, y = pred_UMAP_2), colour = "#de2d26") +
  annotate(geom = 'text', label = 'b', x = -Inf, y = Inf, hjust = -0.5, vjust = 1.5, size = 3)

```{r, out.width="100%", fig.cap = "Predicted 2D embedding overlaid with UMAP embedding of S-curve training data: (a) prediction of 2D embeddings for S-curve training data, and (b) prediction of 2D embeddings for S-curve test data. Once the high-D model is generated, first, for the new data points the nearest high-D mappings of hexagonal bin centroids find by computing Euclidean distance. Then, mapping back the 2D hexagonal bin centroids gives the prediction for the new data points."}

pred_points_training + pred_points_test +
  plot_layout(ncol = 2) &
  theme(plot.tag = element_text(size = 8))
```

@@

```{r, out.width="100%", fig.cap="Visualization of model generated for UMAP embedding for S-curve data in 2D space: (a) triangular mesh, (b), triangular mesh colored by edge type, and (c) triangular mesh after removing the long edges. The edges which has the length greater than $0.5$ are assigned as long edges. To obtain smooth representation in 2D space, the long edges are removed."}

trimesh_s_curve + trimesh_colored_s_curve_umap + trimesh_removed_s_curve_umap +
  plot_layout(ncol = 3) &
  theme(plot.tag = element_text(size = 8))
```

@@

<!--need to add screenshots of the frames and add the link of the video-->
```{r}
tour
```

@@

### 2024-04-01 paper-quollr.tex
Add prediction plots for training and test data

\includegraphics{paper-quollr_files/figure-latex/unnamed-chunk-1-1.pdf}

@@

@@

@@

@@

\texttt{predict\_emb()} function is used to predict 2D embedding for a new data point using the fitted model. This function is useful to predict 2D embedding irrespective of the NLDR technique.

@@

\includegraphics{paper-quollr_files/figure-latex/unnamed-chunk-3-1.pdf}

\includegraphics{paper-quollr_files/figure-latex/unnamed-chunk-4-1.pdf}

\includegraphics{paper-quollr_files/figure-latex/unnamed-chunk-7-1.pdf} \includegraphics{paper-quollr_files/figure-latex/unnamed-chunk-7-2.pdf}

\includegraphics{paper-quollr_files/figure-latex/unnamed-chunk-8-1.pdf} \includegraphics{paper-quollr_files/figure-latex/unnamed-chunk-8-2.pdf}

>>>>

@@

\begin{figure}
\includegraphics[width=1\linewidth]{paper-quollr_files/figure-latex/unnamed-chunk-2-1} \caption{Illustration of key steps of the hexagonal bining algorithm: (a) first, all hexagonal bin centroids are created; (b) second, full hexagon grid is created; and (c) finally, assign the data to hexagons. The centroids are used to generate hexagonal coordinates. Once the full hexagonal grid is generated, by computing the 2D euclidean distance from each NLDR points to hexagonal bin centroids, the nearest NLDR points are allocated to the nearest hexagoanl bin.}\label{fig:unnamed-chunk-2}
\end{figure}

@@

\texttt{predict\_emb()} function is used to predict 2D embedding for a new \(p\)-D data point using the fitted model. This function is useful to predict 2D embedding irrespective of the NLDR technique.

\begin{figure}
\includegraphics[width=1\linewidth]{paper-quollr_files/figure-latex/unnamed-chunk-4-1} \caption{Predicted 2D embedding overlaid with S-curve training UMAP data: (a) Prediction of 2D embeddings for S-curve training data, and (b) Prediction of 2D embeddings for S-curve test data. Once the high-D model is generated, first, for the new data points the nearest high-D mappings of hexagonal bin centroids find by computing Euclidean distance. Then, mapping back the 2D hexagonal bin centroids gives the prediction for the new data points.}\label{fig:unnamed-chunk-4}
\end{figure}

@@

\begin{figure}
\includegraphics[width=1\linewidth]{paper-quollr_files/figure-latex/unnamed-chunk-5-1} \caption{Visualization of model generated for UMAP embedding for S-curve data in 2D space: (a) triangular mesh, (b), triangular mesh colored by edge type, and (c) triangular mesh after removing the long edges. The edges which has the length greater than $0.5$ are assigned as long edges. To obtain smooth representation in 2D space, the long edges are removed.}\label{fig:unnamed-chunk-5}
\end{figure}

@@

@@

\includegraphics{paper-quollr_files/figure-latex/unnamed-chunk-8-1.pdf}

\includegraphics{paper-quollr_files/figure-latex/unnamed-chunk-9-1.pdf}

\includegraphics{paper-quollr_files/figure-latex/unnamed-chunk-11-1.pdf}

\includegraphics{paper-quollr_files/figure-latex/unnamed-chunk-12-1.pdf} \includegraphics{paper-quollr_files/figure-latex/unnamed-chunk-12-2.pdf}

\includegraphics{paper-quollr_files/figure-latex/unnamed-chunk-13-1.pdf} \includegraphics{paper-quollr_files/figure-latex/unnamed-chunk-13-2.pdf}
